---
title: "London School of Economics Coffee Farmer Project Sample Code"
output: pdf_document
date: "2023-09-22"
---

## Summary

In this sample code, I clean and wrangle two panel datasets from the International Coffee Organization. The datasets in question consist
of the domestic consumption and domestic exports of coffee for several countries in the Global South between 1990 and 2018.  

```{r, echo=FALSE}
# Load necessary libraries
  library(dplyr)
  library(tidyverse)
  library(gplots)
  library(ggplot2)
  library(reshape)
  library(fastDummies)
  library(sjmisc)


# Set working directory
  setwd("/Users/ignaciogutierrez/Desktop")
```


## Data Cleaning: Exports Dataset

```{r, echo=FALSE}
# Import datafile 
  dataExp = read.csv("exports-coffee.csv")


# Homogenize strings and remove excess spacing
  dataExp$country <- trimws(dataExp$country)

  
# Replace variable names with year
  col_names <- names(dataExp)
  for (i in 2:length(col_names)) {
    col_names[i] <- str_replace_all(col_names[i], "^X", "yr")
  }
  names(dataExp) <- col_names
  
  
# Correct misspelled strings
  unique(dataExp$country)
  
  dataExp$country[dataExp$country == "Jamaicaaa"] <- "Jamaica"
  dataExp$country[dataExp$country == "Bolivia (Plurinational State of)"] <- "Bolivia"
  dataExp$country[dataExp$country == "Timor-Leste"] <- "Timor Leste"
  dataExp$country[dataExp$country == "Viet Nam"] <- "Vietnam"

  
# Reshape panel data into long form by country-year
  dataExp_long <- dataExp %>%
    pivot_longer(cols = c(yr1990:yr2018), names_to = "year", values_to = "exports")
  
  library(stringr)
  dataExp_long$year <- str_remove_all(dataExp_long$year, "^yr")
```


## Data Cleaning: Domestic Consumption Dataset

```{r, echo=FALSE}
# Import datafile 
  dataDom = read.csv("domestic-consumption-coffee.csv")


# Replace variable names with years
  col_names <- names(dataDom)
  for (i in 3:length(col_names)) {
    col_names[i] <- str_replace_all(col_names[i], "^X", "yr")
  }
  names(dataDom) <- col_names  


# Homogenizing strings: capitalizing country names
  dataDom$country <- str_to_title(dataDom$country)

  
# Check for duplicates and remove them
  dataDom <- dataDom %>%
    arrange(country) %>%
    distinct(country, .keep_all = TRUE)
  print(dataDom$country)
  

# Correct for misspelled strings
  dataDom %>% select(c(country, region))
  print(dataDom)
  
  dataDom <- dataDom %>%
  mutate(
    region = ifelse(
      row_number() %in% c(10, 11, 28),
      c("Central America", "Carribean", "Carribean"),
      region
    )
  )
  print(dataDom)
  

# Reshape panel data into long form by country-year
    dataDom_long <- dataDom %>%
    pivot_longer(cols = c(yr1990:yr2018), names_to = "year", values_to = "consumption")
  
  library(stringr)
  dataDom_long$year <- str_remove_all(dataDom_long$year, "^yr")
```


## Merging Datasets: Exports & Domestic Consumption

```{r, echo=FALSE}
# Merge
  clean_panel <- merge(dataExp_long, dataDom_long, by = c("country", "year"))
  print(clean_panel)
  clean_panel <- clean_panel %>% relocate(region, .before = year)

  
# Create ID number for each country-year observation
  clean_panel <- clean_panel %>%
    mutate(id = factor(country)) %>%
    arrange(id)
  
  clean_panel$id <- as.factor(clean_panel$id)
  clean_panel$id <- as.numeric(clean_panel$id)
  
  clean_panel <- clean_panel %>% relocate(id, .before = country)
    
  
# Conduct balance test
  install.packages("plm")
  library(plm)
  
  my_pdata <- pdata.frame(clean_panel, index = c("id", "year"))
  is.pbalanced(my_pdata)
  unbalancedness <- punbalancedness(my_pdata)
  unbalancedness
```


## Data Visualization

```{r, echo=FALSE}
# Identify countries in the top quartile of avg. export quantity
  clean_panel <- clean_panel %>%
  group_by(country) %>%
  mutate(avg_export = mean(exports))

  quartiles <- quantile(clean_panel$avg_export, probs = c(0, 0.25, 0.50, 0.75, 1))
  clean_panel$quartiles <- cut(clean_panel$avg_export, breaks = quartiles, labels = c(1, 2, 3, 4))
  
  table(clean_panel$country[clean_panel$quartiles == 4])
  
  ## create new data frame for top quartile countries
  filter_clean_panel <- clean_panel %>%
    mutate(keep = quartiles == 4) %>%
    filter(keep)
  print(clean_panel)
  
  
# Generate figures of countries in top quartile
  
  ## create new variable for average consumption in top quartile countries
    filter_clean_panel <- filter_clean_panel %>%
    group_by(country) %>%
    mutate(avg_consumption = mean(consumption))  
    filter_clean_panel <- filter_clean_panel %>% relocate(avg_consumption, .before = quartiles)
  
  ## graph avg. coffee exports for top quartile countries
    ggplot(filter_clean_panel, aes(x = avg_export, y = country)) +
    geom_bar(stat = "identity", position = "dodge", fill = "blue", width = 0.3) +
    labs(title = "Coffee Statistics 1990 - 2018",
         y = "Country",
         x = "Average Exports") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

  ## graph avg. coffee consumption for top quartile countries
    ggplot(filter_clean_panel, aes(x = avg_consumption, y = country)) +
    geom_bar(stat = "identity", position = "dodge", fill = "red", width = 0.3) +
    labs(title = "Coffee Statistics 1990 - 2018",
         y = "Country",
         x = "Average Consumption") +
    theme_minimal() + 
    theme(plot.title = element_text(hjust = 0.5))
```


## Generate New Variables

In this section, I generate two new variables: Total Production (the sum of a country's coffee exports and its domestic consumption of 

coffee) and Percentage Exports (the share of coffee exports from total production).

```{r, echo=FALSE}
# Compute new variables from old
  clean_panel$total_production <- clean_panel$exports + clean_panel$consumption
  clean_panel$pct_exports <- clean_panel$exports / clean_panel$total_production
  print(clean_panel)

  
# Round quantities
  clean_panel <- clean_panel %>%
  mutate(
    exports = round(exports, 3),
    avg_export = round(avg_export, 3),
    total_production = round(total_production, 3),
    pct_exports = round(pct_exports, 3)
  )
```


## Regression

I employ a difference-in-difference specification strategy to assess the effect of the Rwandan government's 2002 national coffee strategy

to boost coffee production by incentivizing investments that shifted production from low-quality to high-quality specialty strains. I 

utilize Burundi's export and domestic consumption data (which was not subject to a national coffee strategy) as the control group in the 

specification strategy, given its geographical similarity and proximity to Rwanda.   

```{r, echo=FALSE}
# Keep only treatment and control countries
  diff_in_diff <- clean_panel[clean_panel$country %in% c("Rwanda", "Burundi"), ]
  print(diff_in_diff)

  
# Generate treatment dummy variable
  diff_in_diff$treatment_Rwanda <- 1
  diff_in_diff$treatment_Rwanda[diff_in_diff$country == "Burundi"] <- 0
  
  
# Generate time dummy for the National Coffee Strategy intervention
  diff_in_diff$post_treatment <- 1
  diff_in_diff$post_treatment[diff_in_diff$year >= 2002] <- 0
  

# Generate diff-in-diff estimator 
  diff_in_diff$estimator <- diff_in_diff$treatment_Rwanda * diff_in_diff$post_treatment
  print(diff_in_diff)
  
  
# Diff-in-diff estimation 
  lm_total_production <- lm(total_production ~ treatment_Rwanda + post_treatment + estimator, data = diff_in_diff)
  summary(lm_total_production)
  
  lm_pct_exports <- lm(pct_exports ~ treatment_Rwanda + post_treatment + estimator, data = diff_in_diff)
  summary(lm_pct_exports)
  

# Data visualization of diff-in-diff results
  
  ## graph for total production (Rwanda vs. Burundi) 
    ggplot(diff_in_diff, aes(x = year, y = total_production)) +
    geom_line(aes(group = country, color = country), linewidth = 1) +
    ylab("Total Production") +
    xlab("Year") +
    theme(legend.title = element_blank()) +
    scale_x_discrete(breaks = seq(1990, 2018, by = 4)) + 
    ggtitle("Diff-in-Diff Estimation: Rwanda vs. Burundi") +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_vline(xintercept = "2002", linetype = "dotted", linewidth = 1)

  ## graph for % share of exports (Rwanda vs. Burundi) 
    ggplot(diff_in_diff, aes(x = year, y = pct_exports)) +
    geom_line(aes(group = country, color = country), linewidth = 1) +
    ylab("% Share of Exports") +
    xlab("Year") +
    theme(legend.title = element_blank()) +
    scale_x_discrete(breaks = seq(1990, 2018, by = 4)) + 
    ggtitle("Diff-in-Diff Estimation: Rwanda vs. Burundi") +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_vline(xintercept = "2002", linetype = "dotted", linewidth = 1)
```


## Conclusion

In conclusion, our data visualization reflects two key insights. The first is the percentage of coffee exports from total production 

became slightly greater and far more consistent for Rwanda following the implementation of its national coffee strategy, particularly when

compared to the same metric and timeframe for Burundi. The second is that total coffee production in Rwanda remained similarly volatile

after 2002 as it was prior to 2002, before plateauing to a more stable level in 2008. Total production levels are roughly similar to 

Burundi's after 2002, although Burundi's show consistent volatility throughout the full timeframe of the analysis. 


