---
title: "London School of Economics Coffee Farmer Project Sample Code"
output: pdf_document
date: "2023-09-22"
---

## Summary

In this sample code, I clean and wrangle two panel datasets from the International Coffee Organization. 
The datasets in question consist of the domestic consumption and domestic exports of coffee for several 
countries in the Global South between 1990 and 2018.

Data Cleaning Process: 
  * (1) Harmonize country names between datasets & standardize variable names
  * (2) Reshape datasets so each row is country-year
  * (3) Merge two datasets together for easier analysis 
  * (4) Run a balance test for merged panel

```{r, echo=FALSE}
# Load necessary libraries
  library(dplyr)
  library(tidyverse)
  library(gplots)
  library(ggplot2)
  library(reshape)
  library(fastDummies)
  library(sjmisc)
  library(stringr)


# Set working directory
  setwd("/Users/ignaciogutierrez/Desktop")
```


## Data Cleaning: Exports Dataset

```{r, echo=FALSE}
# Import datafile
  exports_raw = read.csv("exports-coffee.csv")


# Homogenize strings and remove excess spacing
  exports_raw$country <- trimws(exports_raw$country)


# Replace variable names with year: (Ex. "X2011 -> yr2011")
  col_names <- names(exports_raw)
  for (i in 2:length(col_names)) {
    col_names[i] <- str_replace_all(col_names[i], "^X", "yr")
  }
  names(exports_raw) <- col_names


# Correct misspelled strings
  exports_raw$country[exports_raw$country == "Jamaicaaa"] <- "Jamaica"
  exports_raw$country[exports_raw$country == "Bolivia (Plurinational State of)"] <- "Bolivia"
  exports_raw$country[exports_raw$country == "Timor-Leste"] <- "Timor Leste"
  exports_raw$country[exports_raw$country == "Viet Nam"] <- "Vietnam"


# Reshape panel data into long form by country-year; remove "yr" pre-fix from year observations
  country_exports <- exports_raw %>%
    pivot_longer(cols = c(yr1990:yr2018), names_to = "year", values_to = "exports")

  country_exports$year <- str_remove_all(country_exports$year, "^yr")
```


## Data Cleaning: Domestic Consumption Dataset

```{r, echo=FALSE}
# Import datafile
  consumption_raw = read.csv("domestic-consumption-coffee.csv")


# Replace variable names with years
  col_names <- names(consumption_raw)
  for (i in 3:length(col_names)) {
    col_names[i] <- str_replace_all(col_names[i], "^X", "yr")
  }
  names(consumption_raw) <- col_names


# Homogenizing strings: capitalizing country names
  consumption_raw$country <- str_to_title(consumption_raw$country)


# Check for duplicates and remove them: only 2 of 58 observations are duplicates
  consumption_raw <- consumption_raw %>%
    arrange(country) %>%
    distinct(country, .keep_all = TRUE)


# Correct for misspelled strings
  consumption_raw %>% select(c(country, region))
  print(consumption_raw)
  
  consumption_raw <- consumption_raw %>%
  mutate(
    region = ifelse(
      row_number() %in% c(10, 11, 28),
      c("Central America", "Carribean", "Carribean"),
      region
    )
  )


# Reshape panel data into long form by country-year; remove "yr" pre-fix from year observations
  country_consumption <- consumption_raw %>%
    pivot_longer(cols = c(yr1990:yr2018), names_to = "year", values_to = "consumption")

  country_consumption$year <- str_remove_all(country_consumption$year, "^yr")
```


## Merging Datasets: Exports & Domestic Consumption

```{r, echo=FALSE}
# Merge
  country_db <- merge(country_exports, country_consumption, by = c("country", "year"))
  print(country_db)
  country_db <- country_db %>% relocate(region, .before = year)


# Create ID number for each country
  country_db <- country_db %>%
    mutate(id = factor(country)) %>%
    arrange(id)

  country_db$id <- as.numeric(country_db$id)
  country_db <- country_db %>% relocate(id, .before = country)


# Balance Test: We are looking for significant differences in the relevant co-
# variates across countries. These differences would prevent a comparative 
# analysis among specific countries with similarities which extend beyond the
# data captured here, like proximity, geography, historical events, etc. 
  install.packages("plm")
  library(plm)

  my_pdata <- pdata.frame(country_db, index = c("id", "year"))
  is.pbalanced(my_pdata)
  unbalancedness <- punbalancedness(my_pdata)
  unbalancedness
```


## Data Visualization

```{r, echo=FALSE}
# Identify countries in the top quartile of avg. export quantity
  country_db <- country_db %>%
  group_by(country) %>%
  mutate(avg_export = mean(exports))

  quartiles <- quantile(country_db$avg_export, probs = c(0, 0.25, 0.50, 0.75, 1))
  country_db$quartiles <- cut(country_db$avg_export, breaks = quartiles, labels = c(1, 2, 3, 4))

  table(country_db$country[country_db$quartiles == 4])

  
# Create new data frame for top quartile countries
  top_quartile_panel <- country_db %>%
    mutate(keep = quartiles == 4) %>%
    filter(keep)
  print(country_db)


# Graph average coffee exports and average consumption in top quartile countries
    top_quartile_panel <- top_quartile_panel %>%
    group_by(country) %>%
    mutate(avg_consumption = mean(consumption))
    top_quartile_panel <- top_quartile_panel %>% relocate(avg_consumption, .before = quartiles)

    ggplot(top_quartile_panel, aes(x = avg_export, y = country)) +
    geom_bar(stat = "identity", position = "dodge", fill = "blue", width = 0.3) +
    labs(title = "Coffee Statistics 1990 - 2018",
         y = "Country",
         x = "Average Exports") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))

    ggplot(top_quartile_panel, aes(x = avg_consumption, y = country)) +
    geom_bar(stat = "identity", position = "dodge", fill = "red", width = 0.3) +
    labs(title = "Coffee Statistics 1990 - 2018",
         y = "Country",
         x = "Average Consumption") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
```


## Generate New Variables

In this section, I generate two new variables: Total Production (the sum of a country's 
coffee exports and its domestic consumption of coffee) and Percentage Exports (the share of 
coffee exports from total production).

```{r, echo=FALSE}
# Compute new variables from old
  country_db$total_production <- country_db$exports + country_db$consumption
  country_db$pct_exports <- country_db$exports / country_db$total_production


# Round quantities
  country_db <- country_db %>%
  mutate(
    exports = round(exports, 3),
    avg_export = round(avg_export, 3),
    total_production = round(total_production, 3),
    pct_exports = round(pct_exports, 3)
  )
  print(country_db)
```


## Regression

I employ a difference-in-difference specification strategy to assess the effect of the Rwandan 
government's 2002 national coffee strategy to boost coffee production by incentivizing investments 
that shifted production from low-quality to high-quality specialty strains. I utilize Burundi's 
export and domestic consumption data (which was not subject to a national coffee strategy) as the 
control group in the specification strategy, given its geographical similarity and proximity to 
Rwanda. The data follows country-year observations from 1990 to 2018.

```{r, echo=FALSE}
# Keep only treatment and control countries
  diff_in_diff <- country_db[country_db$country %in% c("Rwanda", "Burundi"), ]
  print(diff_in_diff)


# Generate treatment dummy variable
  diff_in_diff$treat <- 1
  diff_in_diff$treat[diff_in_diff$country == "Burundi"] <- 0


# Generate time dummy for the National Coffee Strategy intervention
  diff_in_diff$post <- 1
  diff_in_diff$post[diff_in_diff$year >= 2002] <- 0


# Generate diff-in-diff estimator
  diff_in_diff$treat_post <- diff_in_diff$treat * diff_in_diff$post
  print(diff_in_diff)


# Diff-in-diff estimation
  lm_total_production <- lm(total_production ~ treat + post + treat_post, data = diff_in_diff)
  summary(lm_total_production)

  lm_pct_exports <- lm(pct_exports ~ treat + post + treat_post, data = diff_in_diff)
  summary(lm_pct_exports)


# Data visualization of diff-in-diff results

  ## graph for total production (Rwanda vs. Burundi)
    ggplot(diff_in_diff, aes(x = year, y = total_production)) +
    geom_line(aes(group = country, color = country), linewidth = 1) +
    ylab("Total Production") +
    xlab("Year") +
    theme(legend.title = element_blank()) +
    scale_x_discrete(breaks = seq(1990, 2018, by = 4)) +
    ggtitle("Diff-in-Diff Estimation: Rwanda vs. Burundi") +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_vline(xintercept = "2002", linetype = "dotted", linewidth = 1)

  ## graph for % share of exports (Rwanda vs. Burundi)
    ggplot(diff_in_diff, aes(x = year, y = pct_exports)) +
    geom_line(aes(group = country, color = country), linewidth = 1) +
    ylab("% Share of Exports") +
    xlab("Year") +
    theme(legend.title = element_blank()) +
    scale_x_discrete(breaks = seq(1990, 2018, by = 4)) +
    ggtitle("Diff-in-Diff Estimation: Rwanda vs. Burundi") +
    theme(plot.title = element_text(hjust = 0.5)) +
    geom_vline(xintercept = "2002", linetype = "dotted", linewidth = 1)
```


## Conclusion

According to the diff-in-diff specification, I find that the Rwandan coffee strategy caused a roughly 
114,000 coffee bag decrease in total production when using Burundi as a comparison control, with 
statistical significance at the 10% level. Furthermore, I also find that the Rwandan coffee strategy 
caused a 0.7% decrease in the share of coffee exports from total production when using Burundi as a 
comparison control, with statistical significance at the 1% level.

Visual inspection shows Burundi and Rwanda demonstrate parallel trends before the policy, with significant 
volatility in coffee production and share of exports. 1994, the year of the Rwandan genocide, proves a 
remarkable exception for Rwanda, with coffee statistics plummeting near zero. Coffee exports increase after 
the strategy's implementation, suggesting that most of the corresponding increase in production is absorbed 
internationally, as was intended by the policy.

Moreover, our data visualization reflects two key insights. Firstly, the share of coffee exports from total 
production became slightly greater and far more consistent for Rwanda following the implementation of the 
national coffee strategy, particularly when compared to the same metric and time frame for Burundi. Secondly, 
total coffee production in Rwanda remained similarly volatile post-2002 as it was pre-2002, before plateauing 
to a more stable level in 2008. Total production levels are roughly similar to Burundi's after 2002, although 
Burundi's demonstrate consistent volatility throughout the full time frame of the analysis. The policy appears 
to have stabilized Rwanda's coffee production industry, permitting the development of a growing export sector.  


